\chapter{Recurgent Field Equation and Lagrangian Mechanics}

\section{Overview}

The principle of stationary action, a cornerstone of modern field theory \autocite{GoldsteinPooleSafko2002, Arnold1989}, governs the dynamics of semantic structures. A single scalar function, the Lagrangian, encodes the interplay of competing semantic forces, and from it the equations of motion derive. This section specifies the Lagrangian for Recurgent Field Theory and derives the Euler-Lagrange field equation for the evolution of coherence across the manifold.

\section{Lagrangian Density}

Semantic dynamics arise from a tension between coherence-seeking flow, the stabilizing influence of attractors, generative autopoietic potential, and regulatory constraints against pathological recursion. The Lagrangian density \(\mathcal{L}\) for a real coherence field \(C_i\) represents these competing influences:
\begin{equation}
\mathcal{L} = \underbrace{\frac{1}{2} g^{ij} (\nabla_i C_k)(\nabla_j C^k)}_{\text{Kinetic Term}} - \underbrace{V(C_{\text{mag}})}_{\text{Potential}} + \underbrace{\Phi(C_{\text{mag}})}_{\text{Autopoiesis}} - \underbrace{\lambda \mathcal{H}[R]}_{\text{Constraint}}
\end{equation}
where summation over repeated indices is implied. The components are:
\begin{itemize}
    \item \textbf{Kinetic Term:} The standard kinetic energy for a multicomponent field, penalizing non-uniform coherence gradients.
    \item \textbf{Potential Term \(V(C_{\text{mag}})\):} A potential function encoding the influence of stable semantic attractors, driving the system toward states of established meaning.
    \item \textbf{Autopoietic Term \(\Phi(C_{\text{mag}})\):} A generative potential active above a critical coherence threshold, driving the formation of novel semantic structures.
    \item \textbf{Humility Constraint \(\mathcal{H}[R]\):} A functional of the recursive coupling tensor \(R\) providing a regulatory mechanism to penalize excessive or unstable recursive amplification. The parameter \(\lambda\) modulates its strength.
\end{itemize}

With this formulation, the resulting field equations are covariant. Any continuous symmetry in the Lagrangian gives rise to a corresponding conservation law, in accordance with Noether's theorem, ensuring the theory respects the fundamental symmetries of theoretical physics \autocite{Noether1918, Lagrange1788, Euler1744, LandauLifshitz1975, PeskinSchroeder1995, Weinberg1995}.

\subsection{Complex Field Formulation}
For systems with wave-like phenomena or phase dynamics, the coherence field must be complex-valued, requiring an extended Lagrangian:
\begin{equation}
\mathcal{L}_{\mathbb{C}} = g^{ij} (\nabla_i C_k)(\nabla_j C^{k*}) - V(|C|) + \Phi(|C|) - \lambda \mathcal{H}[R]
\end{equation}
where \(C^{k*}\) is the complex conjugate of \(C^k\) and \(|C| = \sqrt{g^{ij} C_i C_j^*}\). This formulation, analogous to that of Schrödinger or Dirac fields, models propagating semantic waves and interference effects.

\section{The Principle of Stationary Action}

The action functional, \(S\), is the integral of the Lagrangian density over the semantic manifold \(\mathcal{M}\):
\begin{equation}
S[C_i] = \int_{\mathcal{M}} \mathcal{L}(C_i, \nabla_j C_i, R) \, dV
\end{equation}
where \(dV = \sqrt{|g|} \, d^n p\) is the invariant volume element. The principle of stationary action, \(\delta S = 0\), requires the physical evolution of the field to follow a path toward extremizing this functional.

\section{Euler–Lagrange Field Equation}

The variational principle, applied to the action \(S\), yields the Euler–Lagrange equations for the coherence field \(C_i\) \autocite{Euler1744, Lagrange1788}:
\begin{equation}
\frac{\partial \mathcal{L}}{\partial C_i} - \nabla_j \left( \frac{\partial \mathcal{L}}{\partial (\nabla_j C_i)} \right) = 0
\end{equation}
Substituting the components of \(\mathcal{L}\) gives the explicit equation of motion:
\begin{equation}
\Box C^i + \frac{\partial V(C_{\mathrm{mag}})}{\partial C_i} - \frac{\partial \Phi(C_{\mathrm{mag}})}{\partial C_i} + \lambda \frac{\partial \mathcal{H}[R]}{\partial C_i} = 0
\end{equation}
where \(\Box \equiv g^{jk}\nabla_j \nabla_k\) is the covariant d'Alembertian operator. The potential terms are functions of the coherence magnitude, \(C_{\text{mag}} = \sqrt{g^{ij} C_i C_j}\), and their derivatives are found via the chain rule:
\begin{equation}
\frac{\partial V(C_{\mathrm{mag}})}{\partial C_i} = \frac{dV}{dC_{\mathrm{mag}}} \frac{\partial C_{\mathrm{mag}}}{\partial C_i} = \frac{dV}{dC_{\mathrm{mag}}} \frac{g^{ij} C_j}{C_{\mathrm{mag}}}
\end{equation}
The humility term requires a functional derivative, since \(\mathcal{H}\) depends on the recursive coupling tensor \(R\), which is itself a functional of the underlying semantic field \(\psi\) that generates \(C\):
\begin{equation}
\frac{\partial \mathcal{H}[R]}{\partial C_i(p)} = \int_{\mathcal{M}} \frac{\delta \mathcal{H}[R]}{\delta R_{jkl}(s)} \frac{\delta R_{jkl}(s)}{\delta C_i(p)} \, dV_s
\end{equation}
This term represents a nonlocal feedback loop where the global recursive structure influences local coherence dynamics.

\section{Microscopic Dynamics and Field Coupling}

The Euler-Lagrange equation for \(C_i\) gives the effective dynamics of coherence. However, the theory's axiomatic foundation posits a more fundamental semantic field, \(\psi_i\), from which coherence emerges (\(C_i = \mathcal{F}_i[\psi]\)). A full description of the system must therefore specify the dynamics of \(\psi_i\) and its coupling to \(C_i\).

\subsection{Semantic Field Evolution}

A flow equation describes the evolution of the microscopic field \(\psi_i\):
\begin{equation}
\frac{\partial \psi_i(p, t)}{\partial t} = v_i[\psi, C](p, t)
\end{equation}
The semantic velocity \(v_i\) is driven by gradients in the effective coherence landscape and other recursive forces. A general form for this velocity is:
\begin{equation}
v_i(p, t) = \alpha \cdot \nabla_i C_{\mathrm{mag}}(p, t) + \mathcal{G}_i[\psi](p, t)
\end{equation}
where:
\begin{itemize}
    \item The first term is gradient flow, where \(\psi_i\) evolves to increase local coherence. \(\alpha\) is a coupling constant.
    \item The second term, \(\mathcal{G}_i[\psi]\), includes all other direct recursive forces and influences not mediated by the mean coherence field \(C\). Its specific form depends on the system being modeled.
\end{itemize}
This establishes a bidirectional, multi-scale coupling: microscopic variations in \(\psi_i\) determine the structure of the macroscopic coherence field \(C_i\), that in turn guides the evolution of \(\psi_i\).

\subsection{The Coupled Dynamical System}

The complete theoretical structure comprises a coupled system of partial differential equations:
\begin{enumerate}
    \item \textbf{Microscopic Evolution:} \(\displaystyle \frac{\partial \psi_i}{\partial t} = v_i[\psi, C]\)
    \item \textbf{Macroscopic Definition:} \(C_i = \mathcal{F}_i[\psi]\)
    \item \textbf{Effective Field Equation:} \(\Box C^i + \frac{\partial V}{\partial C_i} - \frac{\partial \Phi}{\partial C_i} + \lambda \frac{\partial \mathcal{H}}{\partial C_i} = 0\)
\end{enumerate}
The system may be solved numerically by iterating between the levels: \(\psi_i\) is updated via its evolution equation, the resulting \(C_i\) is calculated, and \(C_i\) must satisfy the Euler-Lagrange equation. The underlying action principle guarantees the consistency of this procedure, provided the variation \(\delta C_i\) is constrained by admissible variations in \(\psi_i\):
\begin{equation}
\delta C_i(p) = \int_{\mathcal{M}} \frac{\delta C_i(p)}{\delta \psi_j(q)} \, \delta \psi_j(q) \, dV_q
\end{equation}
The dynamics derived from the effective Lagrangian for \(C_i\) therefore remain consistent with the evolution of the fundamental field \(\psi_i\).