\chapter{Formalisms of Agent Communication}
\label{14:formalisms_of_agent_communication}

% ------------------------------------------------------------------------------------------------

\section{Domain Structure and Cross-Domain Mapping}
\label{14.1:domain_structure_and_cross_domain_mapping}

For the purposes of our formalisms, we define a \textit{domain} as a submanifold dedicated to a specific mode of representation, such as the formal domain of mathematics, the lexical domain of language, or the affective domains comprising emotion. Communication, then, is the act of mapping semantic structures from one domain to another. For a mapping to be successful, the underlying logic of the source domain must be translated into the language of the target. Consider the common metaphor \textit{time is money}. For this statement to be meaningful, the semantic structure of economics must map onto the domain of temporality.

In our construction of the Semantic Manifold \(\mathcal{M}\), we can understand it as a collection of these partitioned submanifolds, or domains (\(\mathcal{M} = \bigcup_d \mathcal{M}_d\)), each with its own characteristic metric and organizational principles. Hetero-recursive coupling provides the mechanism for mapping between these distinct semantic spaces. A domain translation tensor, \(T_{\mu\nu}^{(d \to d')}\), formally connects the tangent spaces of different domains, allowing coherence in one to influence another.

The recursive coupling tensor, \(R^\rho_{\mu\nu}\), can thus be decomposed into self-referential (intra-domain) and hetero-referential (inter-domain) components:

\begin{equation}
R^\rho_{\mu\nu}(p, q, t) = R^{\rho, \text{self}}_{\mu\nu}(p, q, t) + R^{\rho, \text{hetero}}_{\mu\nu}(p, q, t)
\end{equation}

where the hetero-recursive part, responsible for cross-domain mapping, is constructed from the latent recursive channel tensor \(\chi^\rho_{\mu\sigma}\) and the domain translation tensor:

\begin{equation}
R^{\rho, \text{hetero}}_{\mu\nu}(p, q, t) = \chi^\rho_{\mu\sigma}(p, q, t) \cdot T_{\ \nu}^{\sigma, (d(q) \to d(p))}
\end{equation}

This provides the fundamental mechanism for inter-agent communication and the construction of meaning across different conceptual frameworks.

% ------------------------------------------------------------------------------------------------

\section{Metaphor and Analogy as Hetero-Recursive Structures}
\label{14.2:metaphor_and_analogy_as_hetero_recursive_structures}

In this formalism, we treat metaphor as more than mere linguistic tool, but expand its description as a fundamental cognitive mechanism which structures understanding by mapping the inferential logic of a concrete source domain onto an abstract target domain. We can draw on the foundational work in conceptual metaphor theory \autocite{LakoffJohnson1980, HofstadterSander2013}. We formalize metaphors and analogies as stable, hetero-recursive mappings between a source domain \(\mathcal{S}\) and a target domain \(\mathcal{T}\). We define a metaphor as a persistent structure in the Semantic Manifold, defined by a set of high-magnitude hetero-recursive couplings:

\begin{equation}
\mathcal{M}_{\mathcal{S} \to \mathcal{T}} = \{(p, q, R^{\rho, \text{hetero}}_{\mu\nu}(p, q, t)) \mid p \in \mathcal{S},\ q \in \mathcal{T},\ \|R^{\rho, \text{hetero}}_{\mu\nu}(p, q, t)\| > \epsilon\}
\end{equation}

The stability of these mappings correspond to the "entrenchment" of a conceptual metaphor, which can be quantified. When such mappings form closed feedback loops, they can give rise to cross-domain amplification and conceptual blending, resulting in genuine semantic innovation. This mechanism is what allows agents to build shared understanding from differing phenomenal perspectives, forming a basis for the emergence of collective intelligence from decentralized interactions \autocite{Surowiecki2004}.

% ------------------------------------------------------------------------------------------------

\section{Inter-Agent Communication Mechanisms}
\label{14.3:inter_agent_communication_mechanisms}

Communication between agents is mediated by these mechanisms:

\textbf{Coherence Broadcast and Reception:}

\begin{equation}
C^{\mu, \mathrm{sent}}(p,t) = \alpha_{\mathcal{A}} \cdot \mathcal{P}_{\mathcal{A}}[C^\mu](p,t)
\end{equation}

\begin{equation}
C^{\mu, \mathrm{received}}(p,t) = \int_{\mathcal{M}} G_{\mathcal{B}}(p,q,t) \cdot C^{\mu, \mathrm{sent}}(q,t) \, dq
\end{equation}

where \(\mathcal{P}_{\mathcal{A}}\) is the projection operator of agent \(\mathcal{A}\) and \(G_{\mathcal{B}}\) is the reception kernel of agent \(\mathcal{B}\).

% ------------------------------------------------------------------------------------------------

\textbf{Semantic Particle Exchange:}

\begin{equation}
\mathcal{C}_{\mathcal{A}} \xrightarrow[\mathrm{geodesic\ path}]{} \mathcal{C}_{\mathcal{B}}
\end{equation}

where concept particles propagate along geodesics between agents.

% ------------------------------------------------------------------------------------------------

\textbf{Recursive Coupling Establishment:}

\begin{equation}
R_{\mu\nu}^{\rho, \mathcal{A},\mathcal{B}}(p, q, t) = \lambda_{\mathrm{com}} \cdot \chi^\rho_{\mu\sigma}(p, q, t) \cdot T_{\ \nu}^{\sigma, (\mathcal{A} \to \mathcal{B})}
\end{equation}

representing direct recursive coupling between agent structures.

% ------------------------------------------------------------------------------------------------

\textbf{Shared Manifold Regions:}

\begin{equation}
\mathcal{S}_{\mathrm{shared}} = \mathcal{A}_{\mathrm{int}} \cap \mathcal{B}_{\mathrm{int}}
\end{equation}

defining common semantic ground.

Communication fidelity is determined by the compatibility of internal structures, metric alignment at interfaces, recursive depth, and wisdom-modulated interpretive accuracy. 