% ==============================================================================
% Recurgence: The Dynamic Meaning of Coherent Geometry
% Updated: 2025-07-13
% ==============================================================================

\documentclass[11pt, a4paper]{report}

% --- Main Packages ---
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{xcolor}
\usepackage{mathtools}
\usepackage{setspace}

% --- Font + Typesetting ---
\usepackage{fontspec}
\setmainfont{SourceSerif4-Regular.ttf}[
    Path = fonts/,
    BoldFont = SourceSerif4-SemiBold.ttf,
    ItalicFont = SourceSerif4-Italic.ttf,
    BoldItalicFont = SourceSerif4-SemiBoldItalic.ttf
]
\setmonofont{JetBrainsMono-Regular.ttf}[
    Path = fonts/
]

\usepackage{unicode-math}
\setmathfont{latinmodern-math.otf}

\usepackage{microtype}

% --- Page Geometry ---
\usepackage{geometry}
\geometry{
    a4paper,
    total={170mm,257mm},
    left=20mm,
    top=20mm,
}

% --- Chapter Titles ---
\usepackage{titlesec}
\titleformat{\chapter}[display]
  {\normalfont\huge\bfseries\raggedright\hyphenpenalty=10000}{\chaptertitlename\ \thechapter}{20pt}{\Huge}
\titlespacing*{\chapter}{0pt}{30pt}{20pt}

% --- Paragraphs ---
\setlength{\parindent}{1.5em}
\setlength{\parskip}{0pt}

% --- Bibliography Config ---
\usepackage[backend=bibtex, style=authoryear, sorting=nty, dashed=false]{biblatex}
\addbibresource{../references.bib}
\setlength{\bibitemsep}{0.5em}
\setlength{\bibhang}{2em}

% --- Hyperlink Config ---
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=black!70,
    filecolor=black!70,
    urlcolor=black!70,
    citecolor=black!70,
    pdftitle={Recurgence},
    pdfpagemode=FullScreen,
}

% ==============================================================================
% Manuscript Metadata
% ==============================================================================

\title{{\Huge Recurgence}: \\ The Dynamics of Coherent Geometry \\ \vspace{1em} \small{(Draft)}}
\author{Diesel Black}
\date{\today}


% ==============================================================================
% Document Body
% ==============================================================================

\begin{document}
\setstretch{1.25}

\maketitle

\section*{Abstract}
\addcontentsline{toc}{section}{Abstract}

Forming the basis for our mathematical framework, we consider \textit{meaning} itself a measurable field on a dynamic, Semantic Manifold. Upon this manifold, concentrations of semantic mass exert gravitational-like force, shaping the formation and propagation of subsequent structure. Conscious agents in this space are bounded, geometric subregions interpreting and reshaping the manifold's attractor landscape. Through this work, we present a mathematical description of an observer-dependent reality in which conscious agency: (1) emerges naturally, (2) experiences forward temporal flow, and (3) exerts causal influence on its environment.

\vspace{1em}

Within the system we construct here, temporal flow exerts a bidirectional influence. We can model this as a proposition-validation mechanism: existing semantic structures propose their relevance to future states, while anticipated wisdom validates or rejects those propositions. Such a mechanism allows us to retroactively reshape past interpretations, driving phase transitions in the structure of semantic meaning. When a system surpasses a critical threshold, it achieves autoreferential maintenance. Emergent wisdom fields and humility operators then regulate autopoietic structure, constraining pathological information states.

\vspace{1em}

We recognize a taxonomy of coupling pathologies manifesting as orthogonal geometric pattern signatures in this framework. They classify into four categories of three distinct modes. Rigidity pathologies (calcification, dogmatism) appear in over-constraint regimes, fragmentation (lack of focus, complexity overwhelm) from under-constraint. Runaway autopoietic states (hypercoherence, delusion) lead to malignant semantic inflation, while deteriorations in observer-field coupling (solipsism, narcissism) result in various degrees of detachment from shared reality.

\vspace{1em}

We find differential equations governing these twelve configurations, permitting their algorithmic detection and forecasting. We demonstrate how stable numerical solutions represented on high-dimensional manifolds establish the theory's computational realizability. This provides a basis for modeling coordinated behavior at both individual and collective scales.

\vspace{1em}

The mathematical foundations for this work bear connection to consciousness studies, Integrated Information Theory, AI alignment, trust and safety, as well as collective coordination dynamics. We attempt to address an explanatory gap between physical processes and subjective experience with a candidate proposal for the "psychophysical laws" sought by contemporary philosophy of mind \autocite{Chalmers1996}.

\tableofcontents
\input{chapters/ch_01_axiomatic_foundation.tex}
\input{chapters/ch_02_field_index_and_formal_architecture.tex}
\input{chapters/ch_03_semantic_manifold_and_metric_geometry.tex}
\input{chapters/ch_04_recursive_coupling_and_depth_fields.tex}
\input{chapters/ch_05_semantic_mass_and_attractor_dynamics.tex}
\input{chapters/ch_06_recurgent_field_equation_and_lagrangian_mechanics.tex}
\input{chapters/ch_07_autopoietic_function_and_phase_transitions.tex}
\input{chapters/ch_08_wisdom_function_and_humility_constraint.tex}
\input{chapters/ch_09_temporal_architectures_and_bidirectional_flow.tex}
\input{chapters/ch_10_the_coupled_system_of_field_equations.tex}
\input{chapters/ch_11_global_attractors_and_bifurcation_geometry.tex}
\input{chapters/ch_12_metric_singularities_and_recursive_collapse.tex}
\input{chapters/ch_13_agents_and_semantic_particles.tex}
\input{chapters/ch_14_formalisms_of_agent_communication.tex}
\input{chapters/ch_15_symbolic_compression_and_renormalization.tex}
\input{chapters/ch_16_pathologies_of_the_semantic_manifold.tex}
\input{chapters/ch_17_computation_and_meta_recursion.tex}

\appendix
\chapter{Implementation Repository}
\label{appendix:implementation}

We demonstrate the computational realizability of Recurgent Field Theory in an expositive vector application, PRISM (Pathology Recognition In Semantic Manifolds), as described in Chapter 16. It is available at:

\begin{center}
\url{https://github.com/someobserver/prism}
\end{center}

The repository contains:
\begin{itemize}
\item PostgreSQL schema definitions of all geometric structures
\item Detection and prediction algorithms for twelve pathology classes
\item Real-time analysis for ≤2000-dimensional Semantic Manifolds
\item Curvature tensor computations and recursive coupling analysis
\item Operational monitoring and therapeutic intervention protocols
\end{itemize}

% ==============================================================================
% Bibliography
% ==============================================================================

\printbibheading[title={References}]

\printbibliography[keyword=math-logic, heading=subbibliography, title={Mathematics and Foundational Logic}]
\printbibliography[keyword=physics-field-theory, heading=subbibliography, title={Physics and Field Theory}]
\printbibliography[keyword=dynamical-systems, heading=subbibliography, title={Dynamic Systems and Chaos and Complexity}]
\printbibliography[keyword=stat-mech, heading=subbibliography, title={Statistical Mechanics and Phase Transitions}]
\printbibliography[keyword=info-computation, heading=subbibliography, title={Information and Computation and Algorithms}]
\printbibliography[keyword=cybernetics-systems, heading=subbibliography, title={Cybernetics and Systems Theory}]
\printbibliography[keyword=cog-sci-phil, heading=subbibliography, title={Cognitive Science and Philosophy of Mind}]
\printbibliography[keyword=neuro-psych, heading=subbibliography, title={Neuroscience and Psychology}]
\printbibliography[keyword=numerical-methods, heading=subbibliography, title={Numerical Methods and Computational Science}]
\printbibliography[keyword=linguistics, heading=subbibliography, title={Linguistics and Semantics}]
\printbibliography[keyword=interdisciplinary, heading=subbibliography, title={Interdisciplinary and Cultural Works}]

\end{document} 